### 2014-03-24 paper/appendix.tex
addressing comments

\section{Proof of Theorem 1}

>>>>

\section{Routing Algorithm}

We want a semi-Eulerian graph (i.e., we want a graph in which every vertex but two have even degree) so that the desired medium can be inserted, traverse every edge, and exit the graph at a different vertex.  Let $G=(V,E_0)$ s.t. $\forall e \in E_0, weight(e)=0, start \in V $ the start point $end \in V$ the end point.  Let $e_{temp} = (start, end)$, $E' = E + e_{temp}$.

We need to connect disconnected subgraphs in $G$.  Let $G_{dis} = \{G_1, G_2, ... G_n\} \in G$ s.t. $\cup{\{G_i \in G_{dis}\}} = G$ and $G_i \cap G_j  = \{ \} \forall G_i, G_j \in G_{dis}, i\neq j$ be disconnected subgraphs in G.  Create $E_{conn} = \{e = (u, v), u \in G_i, degree(u)\%2 = 1, v \in G_j, degree(v) \%2 = 1 i \neq j, weight(e) = distance(u, v)\}$ be the set of all edges connecting odd-degree points in two distinct disconnected subgraphs.  To keep only the most desirable edges in $E_{conn}$, contract each subgraph $G_i$ to a single point $u_i$ with the shortest outgoing edges to each other subgraph: $E_{conn} = \cup_{G_i \in G_{dis}} \{e=(u_i, v), v \in G_j,$ s.t. $\forall E_{conn} \ni e_k = (u, v), u \in G_i, v \in G_j, weight(e_k) \geq weight(e)\}$.  Sort $E_{conn}$ s.t. $E_{conn} = \{e_1, e_2, ... e_n\}, weight(e_1) \leq weight(e_2) \leq \ldots \leq weight(e_n)$.

Let $E_{conn-min} = \{ \}$.  Beginning with $e_1$, add the first edge $e_i \in E_{conn}$ to $E_{conn-min}$.  Then let $E_{dupe} = \{ e = (u, v) \in E_{conn}$ s.t. $u$  is reachable from $v$ along edges $E' \cup E_{conn-min}$.  Let $E_{conn} = E_{conn} \setminus E_{dupe}$.  Repeat until $E_{conn} = \{ \}$.  Let $E' = E' \cup E_{conn-min}$.

At this stage, we want an Eulerian graph: i.e., we need every vertex to be of even degree.  Let $V_{odd} = \{ v \in V$ s.t. $degree(v) \% 2 = 1 \}$.  Create $E_{circuit} = V_{odd} \times V_{odd}$, s.t. $\forall e=(u,v) \in E_{circuit}, weight(e) = distance(u, v)$.  Find minimum matching $E_{circuit-min} \subseteq E_{circuit}$.  Let $E' = E \cup E_{circuit-min}$.

Now to remove our temporary edge and make the graph semi-Eulerian instead of Eulerian, let $E' = E' \setminus e_{temp}$.

### 2014-03-24 paper/exampleobjects.tex
addressing comments

To evaluate our tool's capabilities and to explore new points in the design space, we fabricated a set of five prototype objects designed using the tool.

\tovi{Maybe a materials section 
before this - to talk about the 
specific materilas we used, 
such as the copper paint 
specifications, el lighting, etc.

Need to highlight which aspect of the design space each example object relates to. 
}

\subsection{Touch-sensitive Toys}

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/placeholder/brain.jpg}
\caption{A touch-sensitive brain whose tubes are filled with conductive paint.  Sensing is done on a single wire via SFCS.  Inset shows the internal structure of the tubes generated by our design tool.}
\label{fig:toys}
\end{figure}

@@

We created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in \cite{Harrison-acoustic}.  The brain, rabbit, and boat each have several distinct touch points on their surfaces.  These touch points are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toys}).  We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.  All components were fabricated on a Makerbot Replicator 2.

\subsection{Breathing Bunny}

@@

We created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}).  For this, we used a combination air/vacuum pump: one terminal creates a vacuum while the other creates positive pressure.  Our rabbit has two tubes, one open tube exiting at its mouth and one semi-closed tube capped with rubberlike material in its abdomen.  We connected one tube to each of our pump's terminals, and using a programmable power supply we can mimic a rabbit's breathing pattern.  This example was printed on our Objet Connex 260. \valkyrie{this paragraph is really unclear}

\subsection{Custom Radio}

@@

A custom radio built using tubes allows users to tune to different stations and listen in.  This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (two potentiometers and a speaker) to a microcontroller (in our case, an Arduino).  We hid the microcontroller and a battery-driven power supply in the base of the radio to make it portable.  For the radio functionality, we attached an Si4703 to the Arduino.  We fabricated this device on our Makerbot Replicator 2.

\subsection{Presence-aware Pen Holder}

@@

Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:pens}).  Such information was used by \cite{Mueller-constructable} to determine which physical laser pointer a designer was using to interact with their interface.  Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, one per pen chamber.  Our modification of using a single cable per chamber is possible because we use 6mm diameter fiber optic cables, in comparison to the angel hair tubes used in the original work, and can thus send and receive through the same cable.  At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.   When a pen is in its appointed place, the emitted infrared light is reflected off its bottom and travels back to the receiver, where it registers as bright.  This prototype was built by our Objet Connex 260. \valkyrie{this whole paragraph is unclear}

\subsection{Maze}

@@

We created a maze game with a single particle trapped in a fully enclosed tube.  To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.  We created this prototype on our Makerbot Replicator 2.

\subsection{Animated Neon Sign}

@@

Neon art, perhaps best known for its association with Las Vegas, is traditionally made from hand-formed glass tubes containing neon gas.  The tubes light up when a current is passed through them.  For this type of art, the path of the tubes is of crucial importance, as it determines how the sign will look.  We designed a custom neon sign bearing the UIST logo and the waves of Hawaii.  The piece features open tubes due to support-flushing constraints, however the tubes could also be semi-closed (one end must remain open to connect the EL wire to its power source).  They have been threaded with EL wire which is lit in sequence to create an animation.  Our sign was fabricated on the Objet Connex 260.

>>>>

To evaluate our tool's capabilities and to explore new points in the design space, we fabricated a set of five prototype objects designed using the tool.  All prototypes were fabricated in a single piece, unless described otherwise.

\tovi{Maybe a materials section 
before this - to talk about the 
specific materilas we used, 
such as the copper paint 
specifications, el lighting, etc.
} \valkyrie{I'm not clear on what you are asking for here}

\valkyrie{Bj\"{o}rn, I can't figure out what happened to my figures?  and why they got thrown down into a weird state with my limitations section, which appears now in two places???}

\subsection{Touch-sensitive Toys (open, liquid, star, exterior)}

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/placeholder/brain.jpg}
\caption{A touch-sensitive brain whose tubes are filled with conductive paint.  Sensing is done on a single wire via SFCS.  Inset shows the internal structure of the tubes generated by our design tool.}\input{limitations.tex}

\label{fig:toys}
\end{figure}

@@

We created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in \cite{Harrison-acoustic}.  The brain, rabbit, and boat each have several distinct touch points on their surfaces.  These touch points are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toys}).  We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.  All components were fabricated on a Makerbot, support-free.

\subsection{Breathing Bunny (semi-closed, gas, exterior)}

@@

We created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}).  For this, we used a combination air/vacuum pump: one terminal creates a vacuum while the other creates positive pressure.  Our rabbit has two tubes, one open tube exiting at its mouth and one semi-closed tube capped with rubberlike material in its abdomen.  We connected one tube to each of our pump's terminals, and using a programmable power supply we can mimic a rabbit's breathing pattern.  This example was printed on our Objet in rubberlike material, with support flushed post-print. \valkyrie{this paragraph is really unclear}

\subsection{Custom Radio (open, threadable, exterior)}

@@

A custom radio built using tubes allows users to tune to different stations and listen in.  This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (two potentiometers and a speaker) to a microcontroller (in our case, an Arduino).  We hid the microcontroller and a battery-driven power supply in the base of the radio to make it portable.  For the radio functionality, we attached an Si4703 to the Arduino.  We fabricated this device on our Makerbot, support-free but cut in two halves to allow the Arduino to be inserted.

\subsection{Presence-aware Pen Holder (open, solid, exterior)}

@@

Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:pens}).  Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, one per pen chamber.  We use a single cable per chamber; our 6mm diameter fiber optic cables, in comparison to the angel hair tubes used in the original work, can send and receive through the same cable.  At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.   When a pen is in its appointed place, the emitted infrared light is reflected off its bottom and travels back to the receiver, where it registers as bright.  This prototype was built by our Objet and support was flushed post-print. \valkyrie{this whole paragraph is unclear}

\subsection{Maze (fully enclosed, particulate, tree, interior)}

@@

We created a maze game with a single particle trapped in a fully enclosed tube.  To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.  We created this prototype on our Makerbot in two halves to allow for insertion of the marble.

\subsection{Animated Neon Sign (semi-closed, threadable, interior)}

@@

Neon art, perhaps best known for its association with Las Vegas, is traditionally made from hand-formed glass tubes containing neon gas.  The tubes light up when a current is passed through them.  For this type of art, the path of the tubes is of crucial importance, as it determines how the sign will look.  We designed a custom neon sign bearing the UIST logo and the waves of Hawaii.  The piece features open tubes due to support-flushing constraints, however the tubes could also be semi-closed (one end must remain open to connect the EL wire to its power source).  They have been threaded with EL wire which is lit in sequence to create an animation.  Our sign was fabricated on the Objet in two halves, cutting through the plane of the sign itself, with clear material for the sign around the input design and opaque everywhere else.  We manually flushed support from the non-planar connecting tubes.  The EL wire was threaded post-print, and is controlled by an EL wire sequencer.

### 2014-03-24 paper/io.tex
addressing comments

\section{The Design Space of Pipes}
The design space of pipes is large.  By using different types and topologies of pipes and inserting different media into the completed pipes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.  We describe the  space, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.  We also discuss some possible implications of pipes in 3D prints.

\begin{figure}[h!]
\centering
    \includegraphics[width=3.4in]{figures/tubespace.pdf}
\caption{The design space of pipes.  Pipe types, media, topologies, and design are discussed more fully in the text.}
\label{fig:pipespace}
\end{figure}

\subsection{Types of Pipes}\tovi{Maybe rename types to 
"openings"?}

Pipes can come in four types: open, return, semi-closed, and fully enclosed.  Each of these types offers distinct interaction capabilities (see Figure \ref{fig:pipespace}).  Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.  This includes pipes which function as fountains, but it does not include pipes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).  Pipes whose contents are not physically accessible to the user are ``closed on the user side''.  A pipe which is ``open on the system side'' can have its contents manipulated by a machine, e.g., by resonating, pressurizing, or passing electricity through those contents.  The system does not need to have physical access to the contents in order to manipulate them.

Open pipes originate from the system side and connect the user side, with both ends of the pipe open. This type of pipe may be used to create, for example, capacitive sensors: an open pipe filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the pipe.  Using Swept Frequency Capacitive Sensing \cite{Sato-touche} or other techniques, a user's touch of the open end can be sensed.  An open pipe can also be used for output, for example by creating in-air vortices as in \cite{Sodhi-aireal}.

A return pipe originates at the system and returns back to it.  By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art.  If a return pipe passes very close to the surface of a 3D printed model, warm or cold water passed through the pipe could be used for temperature-based haptic feedback.

Semi-closed pipes are open at the system end (for control of the enclosed medium) and closed at the user end.  The closed interface is interesting when it fabricated from a mobile material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap which can be blown open by air pressure.  Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open pipe's end to behave similarly; another possibility is to use semi-closed pipes as audio-generating resonance chambers.

A fully enclosed pipe has no openings on either the system side or the user side.  Fully enclosed pipes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display). \tovi{used as a container for water 
or other particles that would 
otherwise fall out?}  \valkyrie{I re-worded above.  please check.}  Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.

\subsection{Media in Pipes}

Pipes can be filled with a variety of media to create different interface affordances and capabilities.

``Gas'' comprises all compressible fluids.  Use of fluid pressure inside pipes can create haptic feedback at semi-closed interfaces, or the gases can be used as carriers for scents or fog.  As in \cite{Slyper-pressure}, structures can be engineered to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not pressed), and thus fluid pressure can also be used as an input. \tovi{I realize "fluid" pressure is 
technically correct, but might 
confuse readers, since this is 
the gas pararaph.}

Incompressible fluids (``liquids'') can perform many of the same interface tasks as gases.  One opportunity with liquids is to fill the interior of pipes with them and cap the ends.  In addition, one can use driable conductive fluids, such as copper paint, to coat the interior of pipes and allow them to function as arbitrarily-shaped wires.  This is especially helpful for the creation of a shared ground, or for creating single-wire capacitive interfaces amenable to sensing with SFCS \cite{Sato-touche}.

\tovi{Again, this raises the question of how we are applying the design 
space - is it based on its properties when it is printed, or based on 
any post-print modifications}\valkyrie{re-worded above.  I think we are basing it on status at ``interaction time"}Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.

Particulates, either printed in-place or inserted, can be of varying densities.  A single particle can be used for display.  Sparse particles in a stream of fluid can provide haptic feedback.  Dense particles in a semi-closed pipe allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming}.

Threadable inserted elements, such as electroluminescent (EL) wire or fiberoptic cables, are those that can be threaded through pipes post-printing.  This allows overcoming limitations of printers: for example, a Printed Optics-style interface can be created on an inexpensive consumer-grade 3D printer using pipes and inserted fiber optic cable.

\subsection{Pipe Topologies}

Pipe network topologies enable different types of interactions.

Splitting or mixing pipes offer flexibility in output.  If a maker wished to create a painting device, she might wish to have two system-side pipes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:pipespace}).  Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side pipe, but split the pipe into two (see Figure \ref{fig:pipespace}).\tovi{How about multiple in/
multiple out (in contrast to 
branching/splitting which is 
single in multiple out)
} \tovi{This seciton should refer to the 
possibilities, not to what we 
actually did.}

Star and tree topologies are extensions of the splitting and mixing primitives.  Using a star topology in which the pipes were filled with conductive paint, we created toys with several touch-sensitive areas, see Figure \ref{fig:toys}.

\subsection{Design of Pipes}

Pipe designs may emphasize either their exterior features (connection points) or their interior paths.  These two features lead to different kinds of interfaces.  An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the eyes, ears, tail, etc.  An interface focused on the internal path of the pipe is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the pipe.

\subsubsection{Dimensions of Pipes}

\valkyrie{Reference those pages from Bjoern: http://ttb.com/recent-news/diagrams/ We can talk about wall thickness, tangent length, degree of bend, arg bend angle, inner and outer diameters, centerline radius, etc.}

\subsection{Inputs and Outputs}

\tovi{The table indicates that media is the most
prominent variable in the design space.
Should justify that.
}

@@

\begin{figure*}[t]
\centering
    \includegraphics[width=\textwidth]{figures/designspace.pdf}
\caption{The design space of pipe-based interactions.  Existing systems are written in regular font.  Those created with fabrication are in {\color{blue}blue}. In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated pipes. \tovi{Im not sure how well this table will work. Division between input and output should be more prominent.}}
\label{fig:designspace}
\end{figure*}

\subsubsection{Inputs}

Pipes create opportunities for many types of input sensing across the surface of printed devices.  We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.  We recognize that there are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of pipes.

Touch sensing can be enabled through pipes filled with conductive media.  Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.  Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).  Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.

Pressure can be sensed in multiple ways.  Slyper, et al., in \cite{Slyper-pressure} contribute a set of semi-closed 3D printable primitives that respond to particular actions, like twisting or pushing, with changes in internal air pressure.  These primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes.  Pressure can also be more simply sensed via capacitance using those techniques.

Grasp sensing can also be enabled using Wimmer's FlyEye technique \cite{Wimmer-flyeye}.  This technique involves a pair of optical links at each point of interest: one of each pair is connected to an infrared (IR) LED and the other to a camera.  When some object nears or touches a pair of cables, IR light from the LED is reflected back to the camera and the touched point appears as a bright spot.  Follmer, et al.,'s Jamming User Interfaces \cite{Follmer-jamming} rely on dense particulates enclosed in a flexible volume, and can be sensed optically.  These dense particulates can be printed in-place with desired characteristics (such as varying particle size or shape) using sacrificial support material, or added post-print.  The optical links necessary for sensing with FlyEye or Jamming can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).

Some 3D printers, like our Objet Connex 260, can fabricate rubber-like materials.  Much like Slyper, et al., in \cite{Slyper-shape}, we can sense flexing and bending of prints made on these machines.  While Slyper, et al., created multi-level multi-stage silicone molds with embedded wires to build their sensors, by placing pipes in our rubberlike prints and inserting conductive media afterwards, we can perform the same flex sensing.

Tapping is another input possibility.  Due to differential sound conductivity in different printed materials (in particular, the rubberlike Objet material compared to the hard Objet material \valkyrie{presumably we should get or take some measurements of this and include it in the appendix?}), pipes of a greater conducting material can be embedded in a model of a lesser conducting material.  A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.  Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting pipes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.

@@

\subsubsection{Outputs}

While many input techniques require the use of conductive media inserted into pipes, output media are more varied.  We describe potential pipe-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.  In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible as described above.

\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.  EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.  Colored liquids can be mixed and split before exiting an opaque device, or in a transparent device their mixing, splitting, and paths could be used for display.  PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a pipe could be actuated by a system-side air pump (see Figure \ref{fig:breathe}).  Mechanical motion can be mediated through fluids (a light object which is partially contained in a pipe or located at its end could be pushed by the fluid pressure of the pipe) or threadables (e.g., muscle wire).  Large particles trapped in pipes visible to the user can be used for, e.g., maze games (see Figure \ref{fig:maze}).  We can even direct fluids through a pipe network using a servo-controlled gate, as in Figure \ref{fig:direct}.  Using techniques like those described by Harrison, et al., in \cite{Harrison-buttons}  \valkyrie{this paragraph seems terrible.  I think all these paragraphs might seem terrible...}

\emph{Aural} outputs can be created with gases.  Zoran has successfully 3D printed a functional flute, which in essence is a long open open pipe with semi-closed pipes coverable by keys \cite{Zoran-flute}.  Helmholz resonance (the phenomenon that creates noise when you blow into the top of a glass bottle, which is also how some musical instruments, like ocarinas, work) can be leveraged to create computer-controllable sound chambers.  All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).  Passive sound amplification (like a phonograph horn) and sound redirection are also possible using pipes.

\emph{Haptic} outputs are possible through compressible and incompressible fluids.  Fluids can actuate particulates and create programmable pliability (as discussed above and in \cite{Follmer-jamming}).  Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).  Sodhi, et al., in \cite{Sodhi-aireal} created free-air haptic feedback using controlled vortex generation: this technique could also be reproduced using a series of pipes.  Iwata, et al., explored volumetric fluid-based interactions with programmatically inflatable balloons and an overhead projector in \cite{Iwata-volflex}, and additionally haptic outputs can be used to increase ergonomics, as in the inflatable mouse by Kim, et al. \cite{Kim-inflatablemouse}.

\subsection{Implications}

\tovi{Where does this fit in the design space? This seciton seems out of place.
}

\valkyrie{is this something we want to talk about at all?}

Pipes included in objects can have additional effects beyond their designed intentions.

Pipes included in printed objects allow for their identification.  While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change the physical properties of a printed object.  This includes weight, acoustic resonance, and (if conductive material is present) capacitive signature.

\valkyrie{need to do a rewrite of this that doesn't necessarily focus so deeply on audio only.  we can talk about these other non-visible measurements like weight and capacitive signature.  I also need to actually get an intentional encoding test up and running... we are going to have to be clever about how we stimulate the resonance pipes, though.  using a speaker won't really work: we have to agitate the air in the chambers, so we might need to use something like a reed (as in reed instruments).}

Both identification by recall and intentional encoding are possible.  As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.  Additionally, two objects that are visually identical but which have fully enclosed (or other types) of pipes on their interior can be distinguished acoustically.  Thus we can recall an object's identity once its acoustic signature has been recorded.

\tovi{I would save these details for later, 
when we tlak about our actual 
implementations.}

We have also experimented with intentional encoding.  Semi-closed pipes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed pipe length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).  All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the pipe. \valkyrie{need to provide some guidelines for the number of IDs we can create in this way, and where in the sound spectrum they are.  my sense is that this will depend highly upon the i/o sensitivities of the mic and speaker selected.}

An object's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.  Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.

This is similar to work done by Willis, et al., in \cite{Willis-infrastructs}.  While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.

\valkyrie{in my experiments so far, it seems that the material printed does conduct sound fairly well, so this is feasible.  I think we should get some equipment and do a small test, though, since I don't know how the resonance will change with the big hunk of plastic around the semi-open pipe.  hopefully discussing this with Alex soon.}

>>>>

\section{Tubular Inputs and Outputs}

\valkyrie{What do you guys think about trying to merge this with the example objects section?  It would cut down the text a lot, but it might seem weird to just shunt those in here.}

Once objects have been designed and fabricated, the major flexibility for prototyping with pipes stems from the media inserted in them.  The five types of media--gas, liquid, solid, particulate, and threadable--give rise to unique possibilities in output (visual, aural, tactile/haptic, olfactory/gustatory) as well as input (touch/pressure and others).

@@

\begin{figure*}[t]
\centering
    \includegraphics[width=\textwidth]{figures/designspace.pdf}
\caption{The design space of pipe-based interactions.  Existing systems are written in regular font.  Those created with fabrication are in {\color{blue}blue}. In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated pipes. \tovi{Im not sure how well this table will work. Division between input and output should be more prominent.} \valkyrie{one possibility is making the table larger and adding some short descriptions.  we could throw out a lot of words from the paper that way}}
\label{fig:designspace}
\end{figure*}

\subsection{Outputs}

We describe potential pipe-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.  In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible as described above.

\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.  Pneumatic actuation (as in \cite{Yao-pneui} or \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in \cite{Sodhi-aireal} demo) also offers visual feedback.  Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.  Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.  Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and   EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.

\emph{Aural} outputs can be created with gases.  Helmholz resonance can be leveraged to create computer-controllable sound chambers, as in Zoran's flute \cite{Zoran-flute}.  All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).  Passive sound amplification and sound redirection are also possible using pipes.

\emph{Haptic} outputs are possible through compressible and incompressible fluids.  Fluids can actuate particulates and create programmable pliability \cite{Follmer-jamming}.  Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).  Free-air haptic feedback \cite{Sodhi-aireal} and volumetric interactions \cite{Iwata-volflex} can be pipe-mediated.

\subsection{Inputs}

Pipes create opportunities for many types of input sensing across the surface of printed devices.  We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.  There are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of pipes.

\emph{Touch} sensing can be enabled through pipes filled with conductive media.  Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.  Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toys}).  Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.

\emph{Pressure} can be sensed in multiple ways.  Slyper, et al.,'s primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes \cite{Slyper-pressure}.  Pressure can also be sensed via capacitance.

\emph{Grasp} sensing can be enabled using Wimmer's FlyEye techniques \cite{Wimmer-flyeye} or those described in Jamming User Interfaces \cite{Follmer-jamming}.  The optical links necessary for sensing can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).

Some 3D printers can fabricate rubber-like materials.  Similar to Slyper, et al., in \cite{Slyper-shape}, we can sense \emph{flexing} and \emph{bending} of prints made on these machines: the ability to 3D print devices to sensing bending and flexing significantly cuts molding and assembly time.

\emph{Tapping} is another input possibility.  Due to differential sound conductivity in different printed materials, pipes of a greater conducting material can be embedded in a model of a lesser conducting material.  A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.  Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting pipes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.

@@

\subsection{Fabrication Techniques}

The physical fabrication process used by different machines gives rise to differing design guidelines.  The two machines we used to create our example objects were the Objet Connex 260 and the Makerbot Replicator 2.

The Objet is an inkjet-based 3D printer which jets drops of polymer in very fine layers, curing each layer with an ultraviolet lamp.  It offers a wide range of material types, including clear/opaque, flexible/hard, and black/white.  Additionally, it permits the mixing of two different materials in a single print: a mixed material could include 70\% flexible and 30\% hard, for example, to be firm but somewhat flexible.  The flexible materials can be inflated when they are thin: they elongate 200\% before breaking.  We have found $.05$inches to be a good thickness for inflatable materials.  Although this machine has many extraordinary capabilities, it unfortunately cannot print overhangs of greater than $14^{\circ}$ without support material.  The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface or blasted out with a water jet from the interior.  The support is also soluble in lye.

The Makerbot Replicator 2 is a fused-deposition modeling (FDM) printer, in which we use exclusively Polylactic Acid (PLA) bioplastics.  More material properties are being explored in PLA: there are now transparent and flexible materials.  They cannot be blended like those in the Objet.  The Makerbot has a fairly coarse resolution compared to the Objet, however it permits overhangs up to $45^{\circ}$ without support and also allows bridging (short spans of material that bridge the gap between two structures with no support underneath).  Due to the FDM process it can even recover from greater overhangs and longer bridges, as the material builds up and can reorganize into a structure \valkyrie{how do I say this?  do I need a photo?  print recovery is pretty awesome-looking.}.  The support used by the Makerbot is made of the same material as the model itself, just printed in a different pattern and attached over as small a surface area as possible to ease removal.  Support inside of a pipe may require pliers, or be infeasible to remove dependent upon geometry.

Assembly is easiest on both machines when the model does not require support material.  Models can be ``cut up'' into assembleable pieces to evade support requirements or ease material removal.  We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.

### 2014-03-24 paper/limitations.tex
addressing some comments, but mostly just copying the tovi comments in
addressing comments

The physical fabrication process also leads to some considerations.  To print without support on a hobbyist machine, internal tubes cannot have overhangs of greater than $45^{\circ}$.  On the Objet, support-free prints cannot overhang more than $14^{\circ}$.  Removal of support from printed tubes can be troublesome, especially when their internal geometry and branching structures are complex.

Our design tool is not yet fully capable of realizing all possibilities in the design space of tubes.  It lacks X and Y \valkyrie{probably it will lack the network manipulation to make mixing or splitting tubes...?}.  Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.  We hope to address this in future work.

\valkyrie{what other future work?  miniaturization doesn't seem right, it's possible to use on mobile phones already with their on-board sensors (mic), maybe talking about shape-changing tubes?  post-fabrication modification with drills?  that doesn't sound very convincing.}

>>>>

\valkyrie{Our design tool is not yet fully capable of realizing all possibilities in the design space of tubes.  It lacks the network manipulation to make mixing or splitting tubes, unless they are designed in an input vector graphics file.  Future work should also include assisting makers in the fabrication process itself by automatically cutting objects to reduce support material removal issues.  Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing. --------- this paragraph breaks the latex for some reason?}

We are also eager to test our tool with users and get feedback from makers who use it in the wild.
